---
title: "Storm data analysis"
author: "Yaroslav Tsaregorodtsev"
date: "3/31/2020"
output:
  html_document: default
  pdf_document: default
---


## Librarys loading
```{r, librares, message = FALSE}
library(dplyr)
library(data.table)
library(readr)
library(ggplot2)
library(stringr)
library(stringi)
```
## Storm Data description

Storm Data is an official publication of the National Oceanic
and Atmospheric Administration (NOAA) which documents:

a. The occurrence of storms and other significant weather phenomena having
sufficient intensity to cause loss of life, injuries, significant property damage, and/or
disruption to commerce;

b. Rare, unusual, weather phenomena that generate media attention, such as snow
flurries in South Florida or the San Diego coastal area; and

c. Other significant meteorological events, such as record maximum or minimum
temperatures or precipitation that occur in connection with another event. 

## Data Processing

```{r, loading data, message = FALSE}
#Loadind Storm data
data_path <- list.files(pattern = "repdata_data_StormData.csv",
                        recursive = TRUE, ignore.case = TRUE)
data <- read_csv(data_path, 
                 col_types = cols(BGN_DATE = col_date(format = "%m/%d/%Y %H:%M:%S"),
                                  WFO = col_character(),
                                  END_TIME = col_character(),
                                  BGN_AZI = col_character(),
                                  BGN_LOCATI = col_character(),
                                  END_DATE = col_character(),
                                  CROPDMGEXP = col_character(),
                                  ZONENAMES = col_character(),
                                  REMARKS = col_character(),
                                  END_LOCATI = col_character(),
                                  END_AZI = col_character(),
                                  STATEOFFIC = col_character())) %>%
  data.table()

#Loadind event table
event_table_path <- list.files(pattern = "event_table", 
                               recursive = TRUE, ignore.case = TRUE)
event_table <- read_delim(event_table_path, delim = ";") %>%
  data.table()

#Loadind State list (50 States and District of Columbia)
state_list_path <- list.files(pattern = "State_list",                                
                              recursive = TRUE, ignore.case = TRUE)
state_list <- read_delim(state_list_path, delim = ";") %>%
  data.table()

#Grouping data by main event types and classification for the 3 designators: County/Parish; Zone; and Marine
storm_data <- data %>%
  mutate(Event_Name = case_when(str_detect(EVTYPE %>% toupper(), "ASTRONOMICAL LOW TIDE") ~ "ASTRONOMICAL LOW TIDE",
                                str_detect(EVTYPE %>% toupper(), "AVALANCHE") ~ "AVALANCHE",
                                str_detect(EVTYPE %>% toupper(), "BLIZZARD") ~ "BLIZZARD",
                                str_detect(EVTYPE %>% toupper(), "COASTAL FLOOD") ~ "COASTAL FLOOD", 
                                str_detect(EVTYPE %>% toupper(), "COLD/WIND CHILL") ~ "COLD/WIND CHILL", 
                                str_detect(EVTYPE %>% toupper(), "DEBRIS FLOW") ~ "DEBRIS FLOW", 
                                str_detect(EVTYPE %>% toupper(), "DENSE FOG") ~ "DENSE FOG",
                                str_detect(EVTYPE %>% toupper(), "DENSE SMOKE") ~ "DENSE SMOKE",
                                str_detect(EVTYPE %>% toupper(), "DROUGHT") ~ "DROUGHT",
                                str_detect(EVTYPE %>% toupper(), "DUST DEVIL") ~ "DUST DEVIL",
                                str_detect(EVTYPE %>% toupper(), "DUST STORM") ~ "DUST STORM",
                                str_detect(EVTYPE %>% toupper(), "EXCESSIVE HEAT") ~ "EXCESSIVE HEAT",
                                str_detect(EVTYPE %>% toupper(), "EXTREME COLD/WIND CHILL") ~ "EXTREME COLD/WIND CHILL",
                                str_detect(EVTYPE %>% toupper(), "FLASH FLOOD") ~ "FLASH FLOOD",
                                str_detect(EVTYPE %>% toupper(), "FLOOD") ~ "FLOOD",
                                str_detect(EVTYPE %>% toupper(), "FROST/FREEZE") ~ "FROST/FREEZE",
                                str_detect(EVTYPE %>% toupper(), "FUNNEL CLOUD") ~ "FUNNEL CLOUD",
                                str_detect(EVTYPE %>% toupper(), "FREEZING FOG") ~ "FREEZING FOG",
                                str_detect(EVTYPE %>% toupper(), "HAIL") ~ "HAIL",
                                str_detect(EVTYPE %>% toupper(), "HEAT") ~ "HEAT",
                                str_detect(EVTYPE %>% toupper(), "HEAVY RAIN") ~ "HEAVY RAIN",
                                str_detect(EVTYPE %>% toupper(), "HEAVY SNOW") ~ "HEAVY SNOW",
                                str_detect(EVTYPE %>% toupper(), "HIGH SURF") ~ "HIGH SURF",                  
                                str_detect(EVTYPE %>% toupper(), "HIGH WIND") ~ "HIGH WIND", 
                                str_detect(EVTYPE %>% toupper(), "HURRICANE|TYPHOON") ~ "HURRICANE (TYPHOON)",
                                str_detect(EVTYPE %>% toupper(), "ICE STORM") ~ "ICE STORM",
                                str_detect(EVTYPE %>% toupper(), "LAKE-EFFECT SNOW") ~ "LAKE-EFFECT SNOW",
                                str_detect(EVTYPE %>% toupper(), "LAKESHORE FLOOD") ~ "LAKESHORE FLOOD",
                                str_detect(EVTYPE %>% toupper(), "LIGHTNING") ~ "LIGHTNING",
                                str_detect(EVTYPE %>% toupper(), "MARINE HAIL") ~ "MARINE HAIL",
                                str_detect(EVTYPE %>% toupper(), "MARINE HIGH WIND") ~ "MARINE HIGH WIND",
                                str_detect(EVTYPE %>% toupper(), "MARINE STRONG WIND") ~ "MARINE STRONG WIND",
                                str_detect(EVTYPE %>% toupper(), "MARINE THUNDERSTORM WIND") ~ "MARINE THUNDERSTORM WIND",                          
                                str_detect(EVTYPE %>% toupper(), "RIP CURRENT") ~ "RIP CURRENT", 
                                str_detect(EVTYPE %>% toupper(), "SEICHE") ~ "SEICHE", 
                                str_detect(EVTYPE %>% toupper(), "SLEET") ~ "SLEET", 
                                str_detect(EVTYPE %>% toupper(), "SNOW") ~ "SNOW",
                                str_detect(EVTYPE %>% toupper(), "STORM SURGE/TIDE") ~ "STORM SURGE/TIDE", 
                                str_detect(EVTYPE %>% toupper(), "STRONG WIND") ~ "STRONG WIND",
                                str_detect(EVTYPE %>% toupper(), "THUNDERSTORM WIND|TSTM WIND|THUNDERSTORM") ~ "THUNDERSTORM WIND", 
                                str_detect(EVTYPE %>% toupper(), "TORNADO") ~ "TORNADO",
                                str_detect(EVTYPE %>% toupper(), "TROPICAL DEPRESSION") ~ "TROPICAL DEPRESSION",
                                str_detect(EVTYPE %>% toupper(), "TROPICAL STORM") ~ "TROPICAL STORM",
                                str_detect(EVTYPE %>% toupper(), "TSUNAMI") ~ "TSUNAMI",
                                str_detect(EVTYPE %>% toupper(), "VOLCANIC ASH") ~ "VOLCANIC ASH",
                                str_detect(EVTYPE %>% toupper(), "WATERSPOUT") ~ "WATERSPOUT",
                                str_detect(EVTYPE %>% toupper(), "WILDFIRE") ~ "WILDFIRE",
                                str_detect(EVTYPE %>% toupper(), "WINTER STORM") ~ "WINTER STORM",
                                str_detect(EVTYPE %>% toupper(), "WINTER WEATHER") ~ "WINTER WEATHER",                           
                                TRUE ~ "Other")) %>%
  merge(event_table, by = "Event_Name", all.x = TRUE) %>%
  merge(state_list, by.x = "STATE", by.y = "Code", all.x = TRUE)
```
## Analysis

Storm Data ware considered. All state codes converted to state names (states that have a different code from code 50 states + District of Columbia receive empty name). Also, all weather events ware grouped into general types from Storm Data Event Table (see https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf, section 2.1.1) and there are three designators: County/Parish; Zone; and Marine.

Analysis consists of 3 parts:

1. Crop damage and fatalities for all time of observed data and all states.

2. Total level of Crop damage and fatalities for all states.

3. Crop damage and fatalities for all events and designators.

## Calculating
```{r, calculating}
##calculating total Crop Damage and Fatality for the all States


 damage_state_year <- storm_data %>%
   group_by(BGN_DATE, State, Event_Name) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
  arrange(desc(Crop_damage))
 summary(damage_state_year)
 head(damage_state_year, 10)
 
 damage_state <- storm_data %>%
   group_by(State, Event_Name) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   arrange(desc(Crop_damage))
 summary(damage_state)
 head(damage_state, 10)
 
fat_state_year <- storm_data %>%
   group_by(BGN_DATE, State, Event_Name) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   arrange(desc(Fatality))
 summary(fat_state_year)
 head(fat_state_year, 10)

 fat_state <- storm_data %>%
   group_by(State, Event_Name) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   arrange(desc(Fatality))
 summary(fat_state)
 head(fat_state, 10) 

 damage_fat_event_year <- storm_data %>%
   group_by(BGN_DATE, Event_Name) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   filter(Crop_damage > 0, Fatality > 0)
 
 damage_event <- storm_data %>%
   group_by(Event_Name) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   arrange(desc(Crop_damage))
 head(damage_event, 10) 
 
 fat_event <- storm_data %>%
   group_by(Event_Name) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   arrange(desc(Fatality))
 head(fat_event, 10) 
 
 
 damage_designator_year <- storm_data %>%
   mutate(Year = year(BGN_DATE)) %>%
   group_by(Year, Designator) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   arrange(desc(Crop_damage))
 head(damage_designator_year, 10) 
 
 fat_designator_year <- storm_data %>%
   mutate(Year = year(BGN_DATE)) %>%
   group_by(Year, Designator) %>% 
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) %>%
   arrange(desc(Fatality))
 head(fat_designator_year, 10) 
 
 damage_fat_designator <- storm_data %>%
   mutate(Year = year(BGN_DATE)) %>%
   group_by(Year, Designator) %>%
   summarise(Crop_damage = sum(CROPDMG),
             Fatality = sum(FATALITIES)) 
 head(damage_fat_designator) 
```

## Plots

You can also embed plots, for example:

```{r}
g <- ggplot(damage_state_year, aes(BGN_DATE, Fatality, group = as.factor(State),  color = as.factor(State)))
g + geom_line() + xlab("Date") + ggtitle("Fig. 1 Fatality level for all states")

d <- ggplot(fat_state_year, aes(BGN_DATE, Crop_damage, group = as.factor(State),  color = as.factor(State)))
d + geom_line() + xlab("Date") + ggtitle("Fig. 2 Crop damage for all states")

e <- ggplot(damage_fat_event_year, aes(BGN_DATE, Crop_damage, group = as.factor(Event_Name),  color = as.factor(Event_Name)))
e + geom_line() + xlab("Date") + ggtitle("Fig. 3 Crop damage for all events")

f <- ggplot(damage_designator_year, aes(Year, Crop_damage, group = as.factor(Designator),  color = as.factor(Designator)))
f + geom_line() + xlab("Year") + ggtitle("Fig. 4 Crop damage for all Designator")

h <- ggplot(fat_designator_year, aes(Year, Fatality, group = as.factor(Designator),  color = as.factor(Designator)))
h + geom_line() + geom_point() + geom_smooth() + xlab("Year") + ggtitle("Fig. 5 Fatality level for all Designator")
```

## Results

1. Crop damage and fatalities for all time of observed data and all states.
From the table 1 we can see that the event with a biggest Crop damage is TORNADO in Smithville, Mississippi April 27, 2011 (fig. 2). Nebraska state has a total biggest Crop damage from HAIL. The most mortatality event (HEAT) was happened in Chicago, Illinois in The 1995 (fig. 1)

2. Total level of Crop damage and fatalities for all states. The state with a most mortality level is Illinois from HEAT and Nebraska has a maximum Crop damage from HAIL.

3. Crop damage and fatalities for all events and designators. In the total level event with a maximum Crop damage is a HAIL and an event with a maximum mortalities is TORNADO. How we can see from fig. 4 and fig. 5 County/Parish have the biggest Crop damage, but an absolute maximum for Fatality is achieved for a Zone.